What is an Ancestral Recombination Graph?

### 2022-02-21 paper.tex
Rename figure 3, arg-in-pedigree

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
\end{center}
\caption{\label{fig-pedigree-and-arg}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined as arbitrary inheritance
of genomes between diploid (or any ploidy; see text) individuals.
(A) Each individual consists of two genomes, and each genome is
inherited as a (potentially) recombinant mixture of the individual's parent's
genomes. (B) Genomes are modelled as nodes in a graph, and the edges
denote inheritance. Each edge is annotated with genome coordinates
defining the specific region(s) inherited.
(C) This description of inheritance fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or GARG.
Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of a GARG.
The pedigree is composed of a set of individuals who each carry
two \emph{genomes} (we assume diploidy here for concreteness, but there
is no inherent limitation on ploidy).
Each genome is inherited from one of the individual's parents, and may
be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-pedigree-and-arg}A
has two genomes $A$ and $B$,
inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
whereas $B$ was inherited directly from $I_4$'s genome $G$ without
recombination. The pedigree is shown here for concreteness, but
is not a necessary part of the model. We can reason directly about the
genomes without necessarily knowing how they are combined within
individuals. This ``genome-perspective'' is show in
Fig.~\ref{fig-pedigree-and-arg}B, which depicts the same graph
of the ancestry of the sample genomes $A$--$D$. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-pedigree-and-arg}C).

The focus on modelling genomes rather than events has some interesting
and important consequences. Firstly, the graph may contain nodes that
do not correspond to any particular event in the ancestry of the
samples. For example, node $H$ is the direct ancestor of $D$ in
Fig~\ref{fig-pedigree-and-arg}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that $D$ descends directly from $K$ (see section XXX)
but there is no necessity for this from a representational perspective.
This ability can be useful in practice and could be required if,
% Do we need to spell this out?
for example, we have a deeply sequenced pedigree of
haplodiploids. % e.g. Honeybees? Be great to cite an example!

Secondly, the way that we model recombination is subtly different.
Rather than having a single node that represents the \emph{event},
there are three nodes (representing the recombinant offspring and
two parental genomes) and two edges with associated genome coordinate
intervals (representing the precise inheritance patterns).
For example, in Fig~\ref{fig-pedigree-and-arg}, node $A$ is the
recombinant descendent of $E$ and $F$, inheriting positions
FIXME from $E$ and genome positions FIXME from $F$. Any particular
node can be both a recombinant (it is the recombined product of
its parent's genomes) and a common ancestor (the ancestry from
multiple samples meet at this node; see the next section).
[More stuff here: clarify why we assign coordinates to edges
rather than nodes. Basically lay out all the stuff that we can
talk about \emph{without} going in to the details of
ancestry resolution.]

>>>>

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined as arbitrary inheritance
of genomes between diploid (or any ploidy; see text) individuals.
(A) Each individual consists of two genomes, and each genome is
inherited as a (potentially) recombinant mixture of the individual's parent's
genomes. (B) Genomes are modelled as nodes in a graph, and the edges
denote inheritance. Each edge is annotated with genome coordinates
defining the specific region(s) inherited.
(C) This description of inheritance fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or GARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a GARG.
The pedigree is composed of a set of individuals who each carry
two \emph{genomes} (we assume diploidy here for concreteness, but there
is no inherent limitation on ploidy).
Each genome is inherited from one of the individual's parents, and may
be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}A
has two genomes $A$ and $B$,
inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
whereas $B$ was inherited directly from $I_4$'s genome $G$ without
recombination. The pedigree is shown here for concreteness, but
is not a necessary part of the model. We can reason directly about the
genomes without necessarily knowing how they are combined within
individuals. This ``genome-perspective'' is show in
Fig.~\ref{fig-arg-in-pedigree}B, which depicts the same graph
of the ancestry of the sample genomes $A$--$D$. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

The focus on modelling genomes rather than events has some interesting
and important consequences. Firstly, the graph may contain nodes that
do not correspond to any particular event in the ancestry of the
samples. For example, node $H$ is the direct ancestor of $D$ in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that $D$ descends directly from $K$ (see section XXX)
but there is no necessity for this from a representational perspective.
This ability can be useful in practice and could be required if,
% Do we need to spell this out?
for example, we have a deeply sequenced pedigree of
haplodiploids. % e.g. Honeybees? Be great to cite an example!

Secondly, the way that we model recombination is subtly different.
Rather than having a single node that represents the \emph{event},
there are three nodes (representing the recombinant offspring and
two parental genomes) and two edges with associated genome coordinate
intervals (representing the precise inheritance patterns).
For example, in Fig~\ref{fig-arg-in-pedigree}, node $A$ is the
recombinant descendent of $E$ and $F$, inheriting positions
FIXME from $E$ and genome positions FIXME from $F$. Any particular
node can be both a recombinant (it is the recombined product of
its parent's genomes) and a common ancestor (the ancestry from
multiple samples meet at this node; see the next section).
[More stuff here: clarify why we assign coordinates to edges
rather than nodes. Basically lay out all the stuff that we can
talk about \emph{without} going in to the details of
ancestry resolution.]
